---
title: "Final Project - Top 1000 Movies by IMDb Rating"
author: "Joseph G., Barbara S., Rakendra S."
date: "2026-04-08"
output: html_document
toc: true
toc_float: true
code_folding: hide
theme: Sandstone
number_section: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
library(kableExtra)
library(priceR)

set.seed(42)
```

# Introduction

This project analyzes the IMDb Top 1000 Movies data set, obtained from Kaggle, which contains the 1,000 highest-rated movies on IMDb as of the date of collection. The goal is to explore what patterns appear among highly rated movies, including which directors and actors appear most often, how movie ratings compare across decades, and how factors such as genre, run time, votes, and box office gross relate to IMDb ratings.

Rather than only listing plots and tables, this report uses exploratory data analysis to build a story about what makes these top-ranked movies stand out.

# Data & Methods

The data set was imported from a CSV file named IMDb_top_1000.csv. It contains information about highly rated movies, including title, release year, run time, genre, director, actors, IMDb rating, meta score, number of votes, and gross earnings.
Before analyzing the data, several columns needed to be cleaned. The Run time column contained text such as "142 min", so the word "min" was removed and the values were converted to numeric. The Gross column included commas and non-numeric characters, so those were removed before converting gross revenue to numeric. The release year was also converted to numeric so it could be grouped by decade.

```{r}
my_data <- read.csv("imdb_top_1000.csv")
my_data <- my_data %>%
  mutate(
    Runtime = as.numeric(gsub("[^0-9]", "", Runtime)),
    Gross = as.numeric(gsub("[^0-9.]", "", Gross)),
    Released_Year = as.numeric(gsub("[^0-9]", "", Released_Year))
  )
```
  
  After cleaning, the dataset contains r nrow(my_data) movies and r ncol(my_data) variables. Some missing values remain, especially in columns such as gross revenue and meta score, so calculations use na.rm = TRUE when appropriate.

# Exploratory Data Analysis

## Most Common IMDb Ratings

We first looked at the distribution of IMDb ratings to understand how ratings are spread across the top 1000 movies.

```{r}
# We use a binwidth of 0.1 because ratings are close  (8.1, 8.2, etc.)
p_rating_hist <- ggplot(my_data, aes(x = IMDB_Rating)) +
  geom_histogram(binwidth = 0.1, fill = "red", color = "black") +
  labs(
    title = "Distribution of IMDB Ratings",
    x = "IMDB Rating",
    y = "Number of Movies"
  ) +
  theme_minimal()


ggplotly(p_rating_hist)

```


## Which Director and Lead Actor work together most often? 

This analysis identifies the most frequent collaborations between directors and their lead actors in the dataset.

```{r}

top_duos <- my_data %>%
  group_by(Director, Star1) %>%
  summarise(movie_count = n(), .groups = "drop") %>%
  arrange(desc(movie_count)) %>%
  head(5)

p1 <- ggplot(top_duos, aes(
  x = movie_count,
  y = reorder(paste(Director, "&", Star1), movie_count),
  fill = Director
)) +
  geom_col() +
  labs(
    title = "Top 5 Director & Actor Pairings",
    x = "Number of Movies Together",
    y = "Partnership"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p1)
```
The results show that Akira Kurosawa & Toshirō Mifune have the strongest collaboration, appearing together in 7 films, making them the most frequent director–actor pair. This is followed by Martin Scorsese & Robert De Niro, who worked together in 6 films, indicating another well-established partnership.

Interestingly, Charlie Chaplin appears as both director and lead actor in 6 films, reflecting his unique role as a filmmaker who frequently starred in his own work.


## Do movies with the most votes have the highest ratings? 
 
This analysis examines whether movies with more audience votes also receive higher IMDb ratings.

```{r, warning = FALSE, message = FALSE}

p3 <- ggplot(my_data, aes(x = No_of_Votes, y = IMDB_Rating, text = Series_Title)) +
  geom_point(alpha = 0.5, color = "green") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, color = "red") +
  labs(
    title = "Number of Votes and IMDb Rating",
    subtitle = "Relationship between audience popularity and perceived movie quality",
    x = "Number of Votes",
    y = "IMDb Rating",
    color = "Film",
    caption = "Data: IMDb Top 1000 Movies Dataset"
  ) +
  theme_minimal()

ggplotly(p3, tooltip = "text")
```

The scatterplot shows that highly voted films such as The Shawshank Redemption, The Dark Knight, and Inception cluster in the 7–9 rating range, indicating that popular movies are generally well-rated. However, they are not always the highest-rated films.

For example, movies like 12 Angry Men and The Godfather achieve very high ratings despite having relatively fewer votes compared to blockbuster films. These act as exceptions, showing that critically acclaimed films may not always have the largest audience reach.


##Has the "Golden Age" passed?  Average Ratings over the Decades.
Does people rated older movie more or new ones (Rakendra S.)
```{r, warning = FALSE, message = FALSE}
# Cleaning years and grouping by Decade
decade_trends <- my_data %>%
  mutate(Year = as.numeric(gsub("[^0-9]", "", Released_Year))) %>%
  filter(!is.na(Year)) %>%
  mutate(Decade = (Year %/% 10) * 10) %>%
  group_by(Decade) %>%
  summarise(avg_rating = mean(IMDB_Rating, na.rm = TRUE))

# Line graph for trends
p5 <- ggplot(decade_trends, aes(x = Decade, y = avg_rating)) +
  geom_line(size = 1, color = "black") +
  geom_point(size = 2) +
  labs(title = "Average IMDB Rating by Decade", x = "Decade", y = "Average Rating") +
  theme_minimal()

ggplotly(p5)

```


The line plot shows that average ratings fluctuate across decades rather than following a clear upward or downward trend. Some decades appear slightly higher and then it falls, but the differences are relatively small overall as the ratings are very close.

This suggests that there is  strong evidence of a consistent “Golden Age” of cinema in terms of IMDb ratings. Instead, each decade contains a mix of highly rated and lower rated films, and the average is not that much influenced by a few standout movies than a systematic change in quality over time.


## Which 5 Directors have earned the most Total Gross? 

This analysis identifies the directors with the highest total box office earnings across all their films in the dataset.
 
```{r, warning = FALSE, message = FALSE}
director_wealth <- my_data %>%
  mutate(Gross_Clean = as.numeric(gsub(",", "", Gross))) %>%
  filter(!is.na(Gross_Clean)) %>%
  group_by(Director) %>%
  summarise(
    Total_Movies = n(),
    Total_Gross_Millions = sum(Gross_Clean, na.rm = TRUE) / 1000000,
    Average_IMDB = mean(IMDB_Rating, na.rm = TRUE)
  ) %>%
  arrange(desc(Total_Gross_Millions)) %>%
  head(5)

director_wealth %>%
  rename(
    `Total Movies` = Total_Movies,
    `Total Gross (Millions)` = Total_Gross_Millions,
    `Average IMDB Rating` = Average_IMDB
  ) %>%
  knitr::kable(
    caption = "Top 5 Directors by Total Gross Revenue. A small group of directors dominate box office earnings."
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

The results show that Steven Spielberg and Christopher Nolan rank among the top directors by total gross revenue, driven by multiple high-earning films. These directors consistently produce commercially successful movies that contribute significantly to their overall earnings.

Other top performers include Peter Jackson, known for large-scale franchise films, and James Cameron, whose blockbuster productions generate extremely high global revenue.

##
What are the top 5 directors with the the highest movie rating? by Joseph Godinez

While the previous section examined how often directors appear in the data set, frequency alone does not capture quality. The following table ranks the top 5 directors by their average IMDb rating across all their films in the data set, limited to directors with enough movies to produce a meaningful average

```{r, warning = FALSE, message = FALSE}
 top_5_directors <- my_data %>%  
   group_by(Director)%>%         
   summarise(avg_rating = mean(IMDB_Rating, na.rm = TRUE))%>% 
   arrange(desc(avg_rating))%>%  
   head(5)%>% 
mutate(Rank = row_number(), avg_rating = round(avg_rating, 2))%>%
  select(Rank, Director, avg_rating)
top_5_directors%>%
  kable(col.names = c("Rank", "Director", "Avg IMDB Rating"), align = c("c", "l", "c"))%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
      
```

##
Who are the top 5 lead actors among the top imdp movies and how many times they appear in this data set ranking. (Joseph Godinez)

To identify which actors appear most frequently among the highest-rated films, the data set was grouped by lead actor and ranked by number of appearances. The table below shows the top 10 lead actors along with their share of the data set, followed by a bar chart for visual comparison.

```{r, warning = FALSE, message = FALSE}
top_actors <- my_data%>%
  group_by(Star1)%>%
  summarise(Movie_count = n())%>%
  arrange(desc(Movie_count))%>%
  head(10)%>%
  mutate(Rank = row_number(), Percentage = paste0(round(Movie_count / nrow(my_data)*100, 1), "%"))%>%
  select(Rank, Star1, Movie_count, Percentage)
top_actors%>%
  kable(col.names = c("Rank", "Actor", "Movies in Dataser", "% in Dataset"), align = c("c", "l", "c", "c"))%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
  
g <- ggplot(top_actors, aes(x = Star1, y = Movie_count, fill = Star1)) + geom_col() +labs(title = "Top 5 lead actors and how many time they appear in the data set", x = "Lead actor's name", y = "Amount of times they appear in the dataset" )

ggplotly(g)
```

##
What is the Average box office gross by decade and adjusted inflation gross by decade (Joseph Godinez)

Raw box office numbers can be misleading when comparing movies across decades, since a dollar in 1950 is worth much more than a dollar today. To account for this, gross earnings were adjusted for inflation to 2024 USD using the priceR package. The following chart compares the nominal and inflation-adjusted average gross by decade

```{r, warning = FALSE, message = FALSE}

decade_gross <- my_data %>%
  mutate(
    adjusted_gross = adjust_for_inflation(Gross, Released_Year, "US", to_date = 2024),
    Decade = (Released_Year %/% 10) * 10
  ) %>%
  group_by(Decade) %>%
  summarise(
    Average_Gross = mean(Gross, na.rm = TRUE),
    Average_Adjusted_Gross = mean(adjusted_gross, na.rm = TRUE),
    .groups = "drop"
  )

print(decade_gross)
decade_long <- decade_gross%>%
  pivot_longer(
    cols = c(Average_Gross, Average_Adjusted_Gross), 
    names_to = "Type",
    values_to = "Gross"
  )%>%
  mutate(Type =recode(Type, 
                      "Average_Gross" = "Nominal Gross", "Average_Adjusted_Gross" = "Ajusted Gross (2024)" ))

g <- ggplot(decade_long, aes(x= Decade, y = Gross, colour  =Type))+ geom_line(linewidth = 1.0)+ geom_point(size =3) + labs( title = "Average box office by decade", subtitle = "Nominal vs. Inflation-Adjsuted to 2024 USD", x = "Decade", y = "Average Gross (Millions USD") + theme_minimal()

ggplotly(g)
```
Average box office gross by decade, shown in both nominal and inflation-adjusted 2024 USD. The gap between the two lines widens for older decades, reflecting how significantly inflation erodes raw earnings over time. Adjusted figures reveal that mid-century films were commercially stronger than their nominal numbers suggest.

##
Top 10 Director & Lead Actor Pairs among Top 1000 IMDB Rated Movies (Joseph Godinez)

The following chart shows the ten most frequent director and lead actor pairings in the dataset, based on how many movies they share.
```{r, warning = FALSE, message = FALSE}
pair_data <- my_data %>%
  group_by(Director, Star1) %>%
  summarise(Movie_Count = n(), .groups = "drop") %>%
  slice_max(Movie_Count, n = 10) %>%
  mutate(Pair = paste(Director, "+", Star1))

ggplot(pair_data, aes(x = reorder(Pair, Movie_Count), y = Movie_Count, fill = Movie_Count)) +
  geom_col(width = 0.5) +  geom_text(aes(label = paste(Movie_Count, "films")), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) + labs(title = "Top 10 Director & Lead Actor Pairs",
    subtitle = "Among Top 1000 IMDB Rated Movies",
    x = NULL,
    y = "Number of Movies Together"
  ) +theme_minimal(base_size = 12)

```

##
How does the distribution of ratings compare across decades? (Joseph Godinez)

To examine how ratings vary across time, movies were grouped by decade and visualized using a violin plot. Only decades with at least five movies are included to avoid misleading distributions from very small samples

```{r, warning = FALSE, message = FALSE}
decade_data <- my_data %>%
  mutate(Decade = paste0((Released_Year %/% 10) * 10, "s")) %>%
group_by(Decade) %>%
  filter(n() >= 5)%>%
  ungroup()

ggplot(decade_data, aes(x = Decade, y = IMDB_Rating, fill = Decade)) +
  geom_violin() + stat_summary(fun = mean, geom = "crossbar", width = 0.3, linewidth = 0.5, show.legend = FALSE) + labs(title = "Distribution of IMDB Ratings Across Decades", x = "Decade", y = "IMDB Rating") + theme_minimal(base_size = 13)
```
Distribution of IMDb ratings by decade. Ratings are consistently clustered in the 7.6 – 8.2 range across all decades, with the horizontal bar marking the mean. Older decades show slightly wider distributions, likely reflecting the smaller number of films represented


