---
title: "Project Draft workshop"
author: "Joseph G., Barbara S., Rakendra S."
date: "2026-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r}
library(tidyr)
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(priceR)            

my_data <- read.csv("imdb_top_1000.csv")
glimpse(my_data)

```
<<<<<<< HEAD
# Data Cleaning: emo::ji("smile")

```{r}
my_data$Released_Year <- as.numeric(my_data$Released_Year )
my_data$Gross <- as.numeric(my_data$Gross)
my_data$Runtime <-as.numeric(my_data$Runtime)
```


```{r}
# What are the top 5 directors with the the highest movie rating?

 top_5_directors <- my_data %>%  
   group_by(Director)%>%         
   summarise(avg_rating = mean(IMDB_Rating, na.rm = TRUE))%>% 
   arrange(desc(avg_rating))%>%  
   head(5)                     
 print(top_5_directors)         
p <- ggplot(top_5_directors, aes(x = Director, y = avg_rating, fill = Director)) + geom_col() + labs(title = "TOP 5 DIRECTORS BY AVERGAE RATING", x= "Director's name", y = "IMDB Average rating")       
=======

1. What are the top 5 directors with the the highest movie rating?

```{r}
library(dplyr)
>>>>>>> e2ca0a63e1290033a50de5b6bbc631518800cd67

```
2. What is the relationship between ratings and runtime?

```{r}

my_data <- my_data%>%
  mutate(Runtime = as.numeric(gsub(" min", "", Runtime)))
  
ggplot(my_data, aes(x = Runtime, y = IMDB_Rating)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "People's Rating vs Runtime",
    x = "Runtime",
    y = "IMDb Rating"
  )
```


```{r}
ggplot(my_data, aes(x = Runtime, y = Meta_score)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Critics Score vs Runtime",
    x = "Runtime",
    y = "Meta_score"
  )
```
3. Who are the top 5 lead actors among the top imdp movies and how many times they appear in this data set ranking. 

```{r}
library(dplyr)
top_actors <- my_data%>%
  group_by(Star1)%>%
  summarise(Movie_count = n())%>%
  arrange(desc(Movie_count))%>%
  head(10)
print(top_actors)

library(ggplot2)
g <- ggplot(top_actors, aes(x = Star1, y = Movie_count, fill = Star1)) + geom_col() +labs(title = "Top 5 lead actors and how many time they appear in the data set", x = "Lead actor's name", y = "Amount of times they appear in the dataset" )
library(plotly)
ggplotly(g)
```

4. Which genres tend to have the longest runtime?

```{r}
my_data <- my_data %>%
  mutate(Runtime = as.numeric(gsub("[^0-9]", "", Runtime)))

my_data_genre <- my_data %>%
  separate_rows(Genre, sep = ", ")

genre_runtime <- my_data_genre %>%
  group_by(Genre) %>%
  summarize(avg_runtime = mean(Runtime, na.rm = TRUE)) %>%
  arrange(desc(avg_runtime))

head(genre_runtime)
```


```{r}
ggplot(genre_runtime, aes(x = reorder(Genre, avg_runtime), y = avg_runtime)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Runtime by Genre",
    x = "Genre",
    y = "Average Runtime (minutes)"
  )
```


5. Average gross movies made per decade

```{r}
<<<<<<< HEAD
country <- "US"
inflation_dataframe <- retrieve_inflation_data(country)
countries_dataframe <- show_countries()
=======
library(dplyr)
>>>>>>> e2ca0a63e1290033a50de5b6bbc631518800cd67

decade_gross <- my_data %>%
  mutate(adjusted_gross = adjust_for_inflation(parse_number(Gross), as.numeric(Released_Year), country, to_date = 2024), 
         Decade = (as.numeric(Released_Year) %/% 10) * 10) %>%
  filter(!is.na(Decade), !is.na(Gross)) %>%
  group_by(Decade)
# 
# %>%
#   summarize(
#     Average_Gross = mean(Gross, na.rm = TRUE),
#     Average_adjusted_gross = mean(adjusted_gross, na.rm = TRUE),
#     Movie_Count = n(),
#     .groups = "drop"
#   ) 

decade_gross$Gross
decade_gross$adjusted_gross

ggplot(decade_gross, aes(x = Decade, y = Average_Gross)) + geom_line()+geom_point()
ggplotly(l)
```
 


6. Which genre produces the most highly-rated movies (rating > 8.5)?

```{r}
high_rated <- my_data %>%
  filter(IMDB_Rating > 8.5)

high_rated_genre <- high_rated %>%
  separate_rows(Genre, sep = ", ")

genre_counts <- high_rated_genre %>%
  group_by(Genre) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

head(genre_counts)
```

```{r}
ggplot(genre_counts, aes(x = reorder(Genre, count), y = count)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of Highly Rated Movies (IMDb > 8.5) by Genre",
    x = "Genre",
    y = "Count of Movies"
  )
```

<<<<<<< HEAD
```{r}

```

=======
7. Which Director and Lead Actor work together most often?
```{r}
# Finding the most frequent pairings
top_duos <- my_data %>%
  group_by(Director, Star1) %>%
  summarise(movie_count = n(), .groups = 'drop') %>%
  arrange(desc(movie_count)) %>%
  head(5)

#  Plot
p1 <- ggplot(top_duos, aes(x = movie_count, y = reorder(paste(Director, "&", Star1), movie_count), fill = Director)) +
  geom_col() +
  labs(title = "Top 5 Director & Actor Pairings", x = "Number of Movies Together", y = "Partnership") +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p1)

```


8. Do movies with the most votes have the highest ratings?
```{r}

p3 <- ggplot(my_data, aes(x = No_of_Votes, y = IMDB_Rating, text = Series_Title)) +
  geom_point(alpha = 0.4, color = "black") +
  geom_smooth(method = "lm", color = "red") + # Adds a trend line
  labs(title = "Number of Votes vs. IMDB Rating", x = "Number of Votes", y = "IMDB Rating") +
  theme_light()


ggplotly(p3, tooltip = "text")

```

9. Has the "Golden Age" passed?  Average Ratings over the Decades.
Does people rated older movie more or new ones
```{r}
# Cleaning years and grouping by Decade
decade_trends <- my_data %>%
  mutate(Year = as.numeric(gsub("[^0-9]", "", Released_Year))) %>%
  filter(!is.na(Year)) %>%
  mutate(Decade = (Year %/% 10) * 10) %>%
  group_by(Decade) %>%
  summarise(avg_rating = mean(IMDB_Rating, na.rm = TRUE))

# Line graph for trends
p5 <- ggplot(decade_trends, aes(x = Decade, y = avg_rating)) +
  geom_line(size = 1, color = "black") +
  geom_point(size = 2) +
  labs(title = "Average IMDB Rating by Decade", x = "Decade", y = "Average Rating") +
  theme_minimal()

ggplotly(p5)

```

10. Which 5 Directors have earned the most Total Gross?
```{r}
director_wealth <- my_data %>%
  mutate(Gross_Clean = as.numeric(gsub(",", "", Gross))) %>%
  filter(!is.na(Gross_Clean)) %>%
  group_by(Director) %>%
  summarise(
    Total_Movies = n(),
    Total_Gross_Millions = sum(Gross_Clean, na.rm = TRUE) / 1000000,
    Average_IMDB = mean(IMDB_Rating, na.rm = TRUE)
  ) %>%
  arrange(desc(Total_Gross_Millions)) %>%
  head(5)

print(director_wealth)

```

11. What is the most common IMDB rating?
```{r}
library(ggplot2)
library(plotly)


# We use a binwidth of 0.1 because ratings are close  (8.1, 8.2, etc.)
p_rating_hist <- ggplot(my_data, aes(x = IMDB_Rating)) +
  geom_histogram(binwidth = 0.1, fill = "red", color = "black") +
  labs(
    title = "Distribution of IMDB Ratings",
    x = "IMDB Rating",
    y = "Number of Movies"
  ) +
  theme_minimal()


ggplotly(p_rating_hist)

```
>>>>>>> e2ca0a63e1290033a50de5b6bbc631518800cd67
