---
title: "Project Draft workshop"
author: "Joseph G., Barbara S."
date: "2026-04-08"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(priceR)
library(kableExtra)

my_data <- read.csv("imdb_top_1000.csv")
glimpse(my_data)



my_data$Runtime <-as.numeric(my_data$Runtime)
my_data$Gross <- as.numeric(gsub("[^0-9.]", "", my_data$Gross))
my_data$Released_Year <- as.numeric(my_data$Released_Year)
```

# Data Cleaning:

```{r}
# Converting character variables to numeric variables

my_data$Runtime <-as.numeric(my_data$Runtime)
my_data$Gross <- as.numeric(gsub("[^0-9.]", "", my_data$Gross))
my_data$Released_Year <- as.numeric(my_data$Released_Year)


```

```{r}
# What are the top 5 directors with the the highest movie rating? Joseph Godinez

 top_5_directors <- my_data %>%  
   group_by(Director)%>%         
   summarise(avg_rating = mean(IMDB_Rating, na.rm = TRUE))%>% 
   arrange(desc(avg_rating))%>%  
   head(5)%>% 
mutate(Rank = row_number(), avg_rating = round(avg_rating, 2))%>%
  select(Rank, Director, avg_rating)
top_5_directors%>%
  kable(col.names = c("Rank", "Director", "Avg IMDB Rating"), align = c("c", "l", "c"))%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")%>%
  
 print(top_5_directors)         
      

```

1.  What is the relationship between ratings and runtime?

```{r}

my_data <- my_data%>%
  mutate(Runtime = as.numeric(gsub(" min", "", Runtime)))
  
ggplot(my_data, aes(x = Runtime, y = IMDB_Rating)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "People's Rating vs Runtime",
    x = "Runtime",
    y = "IMDb Rating"
  )
```

```{r}
ggplot(my_data, aes(x = Runtime, y = Meta_score)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Critics Score vs Runtime",
    x = "Runtime",
    y = "Meta_score"
  )
```

```{r}
# Who are the top 5 lead actors among the top imdp movies and how many times they appear in this data set ranking. Joseph Godinez

top_actors <- my_data%>%
  group_by(Star1)%>%
  summarise(Movie_count = n())%>%
  arrange(desc(Movie_count))%>%
  head(10)%>%
  mutate(Rank = row_number(), Percentage = paste0(round(Movie_count / nrow(my_data)*100, 1), "%"))%>%
  select(Rank, Star1, Movie_count, Percentage)
top_actors%>%
  kable(col.names = c("Rank", "Actor", "Movies in Dataser", "% in Dataset"), align = c("c", "l", "c", "c"))%>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")%>%
  

print(top_actors)

g <- ggplot(top_actors, aes(x = Star1, y = Movie_count, fill = Star1)) + geom_col() +labs(title = "Top 5 lead actors and how many time they appear in the data set", x = "Lead actor's name", y = "Amount of times they appear in the dataset" )

ggplotly(g)
```

2.  Which genres tend to have the longest runtime?

```{r}
my_data <- my_data %>%
  mutate(Runtime = as.numeric(gsub("[^0-9]", "", Runtime)))

my_data_genre <- my_data %>%
  separate_rows(Genre, sep = ", ")

genre_runtime <- my_data_genre %>%
  group_by(Genre) %>%
  summarize(avg_runtime = mean(Runtime, na.rm = TRUE)) %>%
  arrange(desc(avg_runtime))

head(genre_runtime)
```

```{r}
ggplot(genre_runtime, aes(x = reorder(Genre, avg_runtime), y = avg_runtime)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Runtime by Genre",
    x = "Genre",
    y = "Average Runtime (minutes)"
  )
```

```{r}

```

3.  Which genre produces the most highly-rated movies (rating \> 8.5)?

```{r}
high_rated <- my_data %>%
  filter(IMDB_Rating > 8.5)

high_rated_genre <- high_rated %>%
  separate_rows(Genre, sep = ", ")

genre_counts <- high_rated_genre %>%
  group_by(Genre) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

head(genre_counts)
```

```{r}
ggplot(genre_counts, aes(x = reorder(Genre, count), y = count)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of Highly Rated Movies (IMDb > 8.5) by Genre",
    x = "Genre",
    y = "Count of Movies"
  )
```

# What is the Average box office gross by decade and adjusted inflation gross by decade (Joseph Godinez)

```{r}

decade_gross <- my_data %>%
  mutate(
    adjusted_gross = adjust_for_inflation(Gross, Released_Year, "US", to_date = 2024),
    Decade = (Released_Year %/% 10) * 10
  ) %>%
  group_by(Decade) %>%
  summarise(
    Average_Gross = mean(Gross, na.rm = TRUE),
    Average_Adjusted_Gross = mean(adjusted_gross, na.rm = TRUE),
    .groups = "drop"
  )
print(decade_gross)

decade_long <- decade_gross%>%
  pivot_longer(
    cols = c(Average_Gross, Average_Adjusted_Gross), 
    names_to = "Type",
    values_to = "Gross"
  )%>%
  mutate(Type =recode(Type, 
                      "Average_Gross" = "Nominal Gross", "Average_Adjusted_Gross" = "Ajusted Gross (2024)" ))

ggplot(decade_long, aes(x= Decade, y = Gross, colour  =Type))+ geom_line(linewidth = 1.0)+ geom_point(size =3) + labs( title = "Average box office by decade", subtitle = "Nominal vs. Inflation-Adjsuted to 2024 USD", x = "Decade", y = "Average Gross (Millions USD",) + theme_minimal()
```

# Top 10 Director & Lead Actor Pairs among Top 1000 IMDB Rated Movies (Joseph Godinez)

```{r}
pair_data <- my_data %>%
  group_by(Director, Star1) %>%
  summarise(Movie_Count = n(), .groups = "drop") %>%
  slice_max(Movie_Count, n = 10) %>%
  mutate(Pair = paste(Director, "+", Star1))

ggplot(pair_data, aes(x = reorder(Pair, Movie_Count), y = Movie_Count, fill = Movie_Count)) +
  geom_col(width = 0.5) +  geom_text(aes(label = paste(Movie_Count, "films")), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) + labs(title = "Top 10 Director & Lead Actor Pairs",
    subtitle = "Among Top 1000 IMDB Rated Movies",
    x = NULL,
    y = "Number of Movies Together"
  ) +theme_minimal(base_size = 12)

```

#How does the distribution of ratings compare across decades? (Joseph Godinez)

```{r}
decade_data <- my_data %>%
  mutate(Decade = paste0((Released_Year %/% 10) * 10, "s"))  # turns 1990 into "1990s"
ggplot(decade_data, aes(x = Decade, y = IMDB_Rating, fill = Decade)) +
  geom_violin()
  labs(title = "Distribution of IMDB Ratings Across Decades", x = "Decade", y = "IMDB Rating") + theme_minimal(base_size = 13)
```
